A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-08-29 paper/paper.tex
Several minor changes after Henry's pass. Ready from my side.

\documentclass[runningheads]{llncs}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,pgfplots}
\usepackage[font=small]{subfig,caption}
\usepackage{algorithm2e}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx}
\graphicspath{{figures/}}
% ==================================================================
% Format tiks and pgfplots
\tikzstyle{densely dashed}=[dash pattern=on 0.75pt off 0.5pt]
\pgfplotsset{ytick style={draw=none}}
\renewcommand{\axisdefaultheight}{185pt}
\pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfplotscreateplotcyclelist{mylist}{%
{black,thick,smooth},
{black!50!white,thick,smooth},
{black,thick,densely dashed,smooth},
{black!50!white,thick,densely dashed,smooth},
{black,double,smooth}
}
\pgfplotsset{
        table/search path={data},
    }
\pgfplotsset{
  /pgfplots/xlabel near ticks/.style={
     /pgfplots/every axis x label/.style={
        at={(ticklabel cs:0.5)},anchor=near ticklabel
     }
  },
  /pgfplots/ylabel near ticks/.style={
     /pgfplots/every axis y label/.style={
        at={(ticklabel cs:0.5)},rotate=90,anchor=near ticklabel}
     }
}
% ==================================================================

\newcommand{\myparagraph}[1]{\smallskip\noindent\textbf{#1}.}
\author{Michael~A.~Bekos, Henry~F\"orster, Christian~Geckeler, Lukas Holl\"ander, Michael~Kaufmann, Amad\"aus~M.~Spallek, Jan~Splett}

@@

\institute{
Wilhelm-Schickhard-Institut f\"ur Informatik, Universit\"at T\"ubingen, Germany\\
\texttt{\{bekos,foersth,geckeler,mk\}@informatik.uni-tuebingen.de}\\
\texttt{\{jan-lukas.hollaender,amadaeus.spallek,jan.splett\} @student.uni-tuebingen.de}
}

%conf for conference version, arxiv for arxiv version
\newcommand{\ver}{arxiv}
\newcommand{\arxapp}[2]{\ifthenelse{\equal{\ver}{conf}}{#2}{#1}}

% ==================================================================
\begin{document}
\maketitle
\linenumbers
% ==================================================================

@@

In Graph Drawing, there exists a really rich literature and a wide range of~techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a difficult and challenging task, mainly due to the~edge crossings that negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics \cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview refer to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

@@

The role of critical vertices is central in our algorithmwiki\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at distance~$i$ from the ones of the vertex-pool (that is the shortest path to any vertex in the vertex-pool has length~$i$)  have higher probability for selection than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Note that this method resembles probabilistic hill climbing approaches. Next, we  describe how to compute its new position in the next drawing.

@@

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
For our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$ (where $w$ and $h$ are the width and the height of the initial drawing, respectively), $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=50$.\footnote{In order to further decrease the time needed for an iteration, we only computed a candidate position for 10 of the 50 rays.}

We compared the quality of the produced drawings based on crossing resolution, total resolution, and angular resolution. Additionally, we consider the aspect ratio, and the number of crossings of the corresponding drawings.

All algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (that is,~the total resolution); see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). Additionally, we evaluate the produced drawings regarding the aspect ratio and the number of crossings. The reported results are on average across different drawings with same number of vertices.

@@

We observe that the restricted variant of our algorithm and the force-directed algorithm of Huang et al.\ yield drawings with comparable number of crossings, which at the same time is significantly smaller than the corresponding number of crossings produced by the two other algorithms; refer to Fig.~\ref{fig:cr-res-3}.

@@

Our primary motivation for this work was our participation to the graph drawing contest in 2017, where we miserably performed\footnote{\texttt{http://www.graphdrawing.de/contest2017/results.html}}; the topic was the maximization of the crossing resolution. Our approach for the contest in 2017 was a mixture of the two force directed algorithms by Argyriou et al. and Huang et al.

@@

\newpage

@@

\begin{figure}[htbp]
\centering
%\subfloat[\label{fig:graph5-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph5-yfiles}
%}

\subfloat[\label{fig:graph5-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-angular}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-total}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph5-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-huang-total}
}
\caption{Drawings of graph 5 from last years graph drawing contest.  
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~Our results, 
(d)--(f)~Results computed with Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, 
(g)--(i)~Results computed with Huang et al~\cite{DBLP:journals/vlc/HuangEHL13}.}
\label{fig:graph5}
\end{figure}

\begin{figure}[htbp]
\centering
%\subfloat[\label{fig:graph9-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph9-yfiles}
%}

\subfloat[\label{fig:graph9-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph9-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-huang-total}
}
\caption{Drawings of graph 9 from last years graph drawing contest. 
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~Our results, 
(d)--(f)~Results computed with Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, 
(g)--(i)~Results computed with Huang et al~\cite{DBLP:journals/vlc/HuangEHL13}.}
\label{fig:graph9}
\end{figure}}{}

>>>>

\documentclass[runningheads]{llncs}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,pgfplots}
\usepackage[font=small]{subfig,caption}
\usepackage{algorithm2e}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx}
\graphicspath{{figures/}}

% ==================================================================
% Format tiks and pgfplots
\tikzstyle{densely dashed}=[dash pattern=on 0.75pt off 0.5pt]
\pgfplotsset{ytick style={draw=none}}
\renewcommand{\axisdefaultheight}{185pt}
\pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfplotscreateplotcyclelist{mylist}{%
{black,thick,smooth},
{black!50!white,thick,smooth},
{black,thick,densely dashed,smooth},
{black!50!white,thick,densely dashed,smooth},
{black,double,smooth}
}
\pgfplotsset{
        table/search path={data},
    }
\pgfplotsset{
  /pgfplots/xlabel near ticks/.style={
     /pgfplots/every axis x label/.style={
        at={(ticklabel cs:0.5)},anchor=near ticklabel
     }
  },
  /pgfplots/ylabel near ticks/.style={
     /pgfplots/every axis y label/.style={
        at={(ticklabel cs:0.5)},rotate=90,anchor=near ticklabel}
     }
}
% ==================================================================
\newcommand{\myparagraph}[1]{\smallskip\noindent\textbf{#1}.}

%conf for conference version, arxiv for arxiv version
\newcommand{\ver}{conf}
\newcommand{\arxapp}[2]{\ifthenelse{\equal{\ver}{conf}}{#2}{#1}}

\author{Michael~A.~Bekos, Henry~F\"orster, Christian~Geckeler, Lukas Holl\"ander, Michael~Kaufmann, Amad\"aus~M.~Spallek, Jan~Splett}

@@

\institute{
Institut f\"ur Informatik, Universit\"at T\"ubingen, T\"ubingen, Germany\\
\texttt{\{bekos,foersth,geckeler,mk\}@informatik.uni-tuebingen.de}\\
\texttt{\{jan-lukas.hollaender,amadaeus.spallek,jan.splett\} @student.uni-tuebingen.de}
}

% ==================================================================
\begin{document}
\maketitle
%\linenumbers
% ==================================================================

@@

In Graph Drawing, there exists a rich literature and a wide range of~techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a difficult and challenging task, mainly due to the~edge crossings that negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics \cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview refer to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

@@

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at graph-distance~$i$ from the ones of the vertex-pool have higher probability for selection than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we describe how to compute its new position in the next drawing. Note that our method resembles probabilistic hill climbing approaches.

@@

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
%We compared the quality of the produced drawings based on the following aesthetic~properties:
%
%\begin{multicols}{2}
%\begin{enumerate}[P.1.]
%\item \label{p:cr-res} crossing resolution
%\item \label{p:to-res} total resolution
%\item \label{p:an-res} angular resolution
%\columnbreak
%\item \label{p:as-rat} aspect ratio
%\item \label{p:no-xing} number of crossings
%\end{enumerate}
%\end{multicols}
%
Since all algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (by maximizing~the total resolution), we adjusted each of the algorithms to maximize exclusively the corresponding measures; see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). The reported results are on average across different drawings with same number of vertices. Finally, we mention that for our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$, where $w$ and $h$ are the width and the height of the initial drawing, respectively, $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=50$; also, in order to further decrease the time needed for an iteration, we only computed a candidate position for 10 of the 50 rays.

%We compared the quality of the produced drawings based on crossing resolution, total resolution, and angular resolution. Additionally, we consider the aspect ratio, and the number of crossings of the corresponding drawings.

@@

Regarding the number of crossings, the restricted variant of our algorithm and the force-directed algorithm of Huang et al.\ yield drawings with comparable number of crossings, which at the same time is significantly smaller than the  number of crossings produced by the two other algorithms; see Fig.~\ref{fig:cr-res-3}.

@@

Our primary motivation for this work was our participation to the Graph Drawing contest in 2017, where we miserably performed\footnote{\texttt{http://www.graphdrawing.de/contest2017/results.html}}; the topic was the maximization of the crossing resolution. Our approach for the contest in 2017 was a mixture of the two force directed algorithms by Argyriou et al. and Huang et al.

@@

%\newpage

@@

In this section, we present drawings of the 5th and of the 9th graph given in the Graph Drawing contest 2017 that are produced by different variants of our algorithm and of the algorithms by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, and by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}; see Figures~\ref{fig:graph5} and~\ref{fig:graph9}, respectively. Each variant was obtained by adjusting each of the aforementioned algorithms to optimize the crossing resolution, the angular resolution and the total resolution, respectively; the aesthetic criterion optimized by each variant is reported in the caption of its corresponding subfigure.

\begin{figure}[h!]
\centering
%\subfloat[\label{fig:graph5-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph5-yfiles}
%}

\subfloat[\label{fig:graph5-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-angular}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-total}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph5-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-huang-total}
}
\caption{Different drawings of the 5th graph given in the Graph Drawing 2017 contest produced by different variants of
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~our algorithm, 
(d)--(f)~the algorithm by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, and
(g)--(i)~the algorithm by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.
The aesthetic criterion optimized by each variant is reported in the caption of its subfigure.}
\label{fig:graph5}
\end{figure}

\begin{figure}[p]
\centering
%\subfloat[\label{fig:graph9-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph9-yfiles}
%}

\subfloat[\label{fig:graph9-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph9-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-huang-total}
}
\caption{Different drawings of the 9th graph given in the Graph Drawing 2017 contest produced by different variants of
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~our algorithm, 
(d)--(f)~the algorithm by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, and
(g)--(i)~the algorithm by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.
Each variant was obtained by optimizing a different aesthetic criterion, which is named in the caption of each subfigure.}
\label{fig:graph9}
\end{figure}}{}
